sdmTMB statistical model description

### 2021-11-30 doc/appendix-model.Rmd
Develop model description

---
title: Detailed model description
output:
  bookdown::pdf_document2:
  toc: false
number_sections: false
fig_caption: true
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
---

@@

$$
\begin{aligned}
  \mathbb{E}(y_{\s,t}) &= \mu_{\s,t},\\
  \mu_{\s,t} &= f^{-1}
  \left( \boldsymbol{X} \boldsymbol{\beta} + \omega_s + \epsilon_{\s,t} \right),\\
  \boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
  \boldsymbol{\epsilon}_t &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
\end{aligned}
$$
where $f$ is a link function (e.g., log or logit) and $f^{-1}$ is its inverse link.

* $\boldsymbol{X} \boldsymbol{\beta}$ is defined by the `formula` argument and
  represents the fixed effect model matrix and a corresponding vector of
  coefficients

* $\omega_s$ are included if `spatial = 'on'`

* $\epsilon_{\s,t}$ are included by default if there are multiple time elements (`time` argument is not `NULL`) but can be set to IID (`IID`; default), AR1 (`AR1`), random walk (`RW`), or off (`'off`) via the `spatiotemporal` argument

# Random or multi-level intercepts

@@

where $\alpha_g$ is an example optional "random" intercept---an intercept
centered on zero that varies by level $g$ and is constrained by
$\sigma_\alpha$. This is defined by the `formula` argument via the `(1 | g)`
syntax as in lme4 or glmmTMB.

# Time-varying regression parameters

@@

# Spatial regression parameters

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \zeta_s x_t + \ldots \right),\\
  \boldsymbol{\zeta} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\zeta \right),
\end{aligned}
$$

where $\zeta_s$ is a regression parameter that varies spatially and $x_t$ is a predictor that is constant for a given time $t$. Currently, $x_t$ must be the time variable itself. If you are interested in using a different predictor, contact the developers.

# AR1 spatiotemporal random fields

Dropping the optional $\omega_s$ for simplicity:

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

where $\rho$ is the correlation between subsequent spatiotemporal random
fields. The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.

# Random walk spatiotemporal random fields

Dropping the optional $\omega_s$ for simplicity:

# Offset terms

@@

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + O_{\s,t} + \ldots \right),
\end{aligned}
$$
where $O_{\s,t}$ is an offset term---a **log transformed** variable without
a coefficient (assuming a log link).

@@

$N = 1$ (i.e., 'size' currently fixed) and $\mu$ is probability

@@

$$
\operatorname{Normal} \left( \mu, \phi \right)
$$
where $\phi$ is the standard deviation (following Stan convention of SD not variance).

@@

## [Negative Binomial](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)

Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c)

@@

Variance scales quadratically with mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$.

@@

@@

<!-- EW: Because of space limits, I might suggest moving all this to the SI -->
  The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as $\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1}
  (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right)$,
  <!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
    where $\tau^2$ represents the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents
  a scaling parameter that is estimated (REF).
  The parameter $\nu$ controls the smoothness of the covariance function.
  We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].

>>>>

---
title: sdmTMB statistical model description
output:
  bookdown::pdf_document2:
    toc: false
number_sections: false
fig_caption: true
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
---

@@

As described in the main paper, the complete sdmTMB model can be written as:

$$
\begin{aligned}
\mathbb{E}(y_{\s,t}) &= \mu_{\s,t},\\
\mu_{\bm{s},t} &=
f^{-1} \left( \bm{X}_{1,\bm{s},t} \bm{\beta} +
\bm{\alpha}_g +
\bm{X}_{2,\bm{s},t} \gamma_t +
\bm{X}_{3,t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right),
\end{aligned}
$$

where

* $y$ represents the response data;
* $\mu$ represents the mean;
* $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
* $\bm{X}_{1,\bm{s},t}$, $\bm{X}_{2,\bm{s},t}$, and $\bm{X}_{3,\bm{s},t}$ represent design matrices;
* $\bm{\beta}$ represents a vector of fixed-effect coefficients;
* $\alpha_{g}$ represents random intercepts by group $g$, $\alpha_{g}\sim \mathrm{N}(0,\sigma^2_\alpha)$;
* $\gamma_{t}$ represents time-varying coefficients (a random walk), $\gamma_{t} \sim \mathrm{N}(\gamma_{t-1},\sigma^2_\gamma)$;
* $\zeta_{\bm{s}}$ represents spatially varying coefficients (a random field), $\zeta_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\zeta)$;
* $\omega_{\bm{s}}$ represents a random spatial component (a random field), $\omega_{\bm{s}} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_\omega)$; and
* $\epsilon_{\bm{s},t}$ represents a spatiotemporal component (a random field), $\epsilon_{\bm{s},t} \sim \mathrm{MVN}(\bm{0},\bm{\Sigma}_{\epsilon})$.

We now split this model apart to describe the various parts in more detail

## Fixed- or main-effect portion:

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} \ldots \right)
\end{aligned}
$$

Within `sdmTMB()`,

* $\boldsymbol{X} \boldsymbol{\beta}$ is defined by the `formula` argument and
  represents the fixed effect model matrix and a corresponding vector of
  coefficients.

* Get into smoothers, breakpoints, etc. here

## Spatial random fields

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \omega_s + \ldots \right),\\
\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\ 
\end{aligned}
$$

* $\omega_s$ are included if `spatial = 'on'` and omitted if `spatial = 'off'`.

## Spatiotemporal random fields

Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) but can be set to IID (`'iid'`; default), AR1 (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.

### IID spatiotemporal random fields

$$
\begin{aligned}
\mu_{\bm{s},t} &= f^{-1} \left( \ldots + \epsilon_{\s,t} + \ldots \right),\\
\boldsymbol{\epsilon}_t &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right).
\end{aligned}
$$

* $\epsilon_{\s,t}$ are \ldots

### AR1 spatiotemporal random fields

Dropping other components for simplicity:

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \boldsymbol{X} \boldsymbol{\beta} + \delta_{\s,t} \right),\\
\boldsymbol{\delta}_{t=1} &\sim \operatorname{MVNormal} (\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon}),\\
\boldsymbol{\delta}_{t>1} &= \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon}_t,  \:
\boldsymbol{\epsilon}_t \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields. The $\rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2}$ scales the spatiotemporal variance by the correlation such that it represents the steady-state marginal variance.

### Random walk spatiotemporal random fields

Dropping other components for simplicity:

<!-- * $\alpha_g$, are defined by an lme4-style formula, e.g. `formula = y ~ x + (1 | g)`, where `g` represents a column with grouping factors. -->

<!-- * $\gamma_{t}$, time-varying coefficients, are defined by the `time_varying` formula argument. -->

<!-- * $\zeta_{\bm{s}}$, IID random intercepts, are defined by the `spatial_varying` formula argument. -->

<!-- * $f$ and $f^-1$ functions are defined by the `family` argument. E.g., `family = Gamma(link = "log")`. -->

## Spatially varying coefficients

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \bm{X}_{3,t} \zeta_{\bm{s}} + \ldots \right),\\
  \boldsymbol{\zeta} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\zeta \right),
\end{aligned}
$$

where $\boldsymbol{\zeta}$ is a random field representing a spatially varying coefficient. Usually, $\bm{X}_{3,t}$ would represent a predictor that is constant for a given time $t$. Currently, $\bm{X}_{3,t}$ is restricted to a single covariate, supplied by a formula to `spatial_varying`.

## IID random or multi-level intercepts

@@

where $\alpha_g$ is an example optional "random" intercept---an intercept
centered on zero that varies by level $g$ and is constrained by
$\sigma_\alpha$. This is defined by the `formula` argument via the `(1 | g)`
syntax as in lme4 or glmmTMB. There can be multiple random intercepts. 
E.g., `(1 | g) + (1 | h)`, in which case they are assumed independent and uncorrelated from each other.

## Time-varying regression parameters

@@

If `time` is defined in the call to `sdmTMB()`, then `time_varying ~ 1` implicitly represents a time-varying intercept.

## Offset terms

@@

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + O_{\s,t} + \ldots \right),
\end{aligned}
$$

where $O_{\s,t}$ is an offset term---a **log transformed** variable without
a coefficient (assuming a log link).

@@

where $N$ is the size or number of trials and $\mu$ is the probability.

Following the structure of `stats::glm()` and glmmTMB, a binomial family can be specified in one of 4 ways:

1. the response may be a factor (and the model classifies the first level versus all others)
2. the response may be binomial (0/1)
3. the response can be a matrix of form `cbind(success, failure)`, and 
4. the response may be the observed proportions, and the `weights` argument is used to specify the Binomial size ($N$) parameter (`prob ~ ..., weights = N`).

@@

$$
\operatorname{Normal} \left( \mu, \phi \right)
$$
where $\mu$ is the mean and $\phi$ is the standard deviation (following Stan convention of SD not variance).

@@

## [Negative Binomial 2 (NB2)](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)

@@

Where $\mu$ is the mean and $\phi$ is the dispersion parameter.
The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$.

Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.

## [Negative Binomial 1 (NB1)](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351)

$$
\operatorname{NB1} \left( \mu, \phi \right)
$$
Where $\mu$ is the mean and $\phi$ is the dispersion parameter.
The variance scales linearly with the mean $\mathrm{Var}[y] = \mu + \mu / \phi$.

Internally parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.

@@

where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 1, and $\phi$ is the dispersion parameter.

@@

The Matérn defines the covariance $\Phi \left( s_j, s_k \right)$ between spatial locations $s_j$ and $s_k$ as 
$$
\Phi\left( s_j,s_k \right) = \tau^2/\Gamma(\nu)2^{\nu - 1} (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right),$$
<!-- TODO: paraphrase this to avoid self plagiarizing the synopsis report! -->
where $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma function, $K_\nu$ represents the Bessel function, $d_{jk}$ represents the Euclidean distance between locations $s_j$ and $s_k$, and $\kappa$ represents a scaling parameter that controls the decorrelation distance (REF).
The parameter $\nu$ controls the smoothness of the covariance function.
We set $\nu = 1$, which lets us take advantage of the Stochastic Partial Differential Equation (SPDE) approximation to the GMRF to greatly increase computational efficiency [@lindgren2011].
